---
title: "NOAA Data Cleaning Script"
author: "Danielle D."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

# Introduction

Open source sea surface temperature data (sst) in Celsius for Monterey county. I wasn't able to find sst data for Sonoma County.

Data sourced from NOAA OISST-V2-AVHRR Daily (1/4 degree By time, depth, latitude, longitude), accessed on February 25, 2026. NOAA data are in the public domain.
	
Huang, Boyin; Liu, Chunying; Banzon, Viva F.; Freeman, Eric; Graham, Garrett; Hankins, William; Smith, Thomas M.; Zhang, Huai-Min. (2020): NOAA 0.25-degree Daily Optimum Interpolation Sea Surface Temperature (OISST), Version 2.1. [indicate subset used]. NOAA National Centers for Environmental Information. https://doi.org/10.25921/RE9P-PT57. Accessed [February 2026].

SST has ties to marine heatwaves, upwelling strength, and forage fish 
distribution, all of which could affect arrival for shearwaters. 

# Loading Workspace & Set Up

```{r}
# packages
library(readr) # read data
library(janitor) # standardize column names
library(tidyverse) # data manipulation
library(lubridate) # date handling
library(ggplot2) # plotting
library(sf) # mapping
  library(rnaturalearth) # base maps
```

## Reading in the Data

```{r}
# Load --

# use "here" to manage raw files housed out of project (too large for GitHub)
library(here)

#  define path to raw SST data (outside project)
raw_data_dir <- "~/Desktop/R/Shearwater raw data/"

# load CSV for Monterey County SST
monterey_sst <- read_csv(
  file.path(
    raw_data_dir,
    "Monterey_ncdc_oisst_v2_avhrr_by_time_zlev_lat_lon_3140_bcab_6602.csv"
  ),
  show_col_types = FALSE
)

# check
head(monterey_sst)
```
## Creating Master Dataframes

```{r}
# Clean --

# make remove first row, set date class, make columns numeric
monterey_sst <- monterey_sst %>%            
  slice(-1) %>% # remove additional header data
  mutate( 
    datetime = ymd_hms(time, tz = "UTC"),   # convert character to POSIXct
    date = as_date(datetime),               # extract date
    time = format(datetime, "%H:%M:%S")     # extract time as string
  ) %>%
  mutate(
    sst = as.numeric(sst),
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude),
    anom = as.numeric(anom)
    )

# change longitude degrees 0-360 to the −180–180 system to map
monterey_sst$longitude <- ifelse(
  monterey_sst$longitude > 180,
  monterey_sst$longitude - 360,
  monterey_sst$longitude
)

# check
head(monterey_sst)
```

# Integrity Analysis

```{r}
# Explore --

# 1) date - check range (UTC)
date_range_m <- range(monterey_sst$date)
date_range_m
# ok - date range is 2/28/2020 to 12/26/2025

# 2) time - check range
time_range_m <- range(monterey_sst$time)
time_range_m
# remove - not needed since all sst records taken at 12:00 PM

# 3) depth (m) - check range 
depth_range_m <- range(monterey_sst$depth, na.rm = TRUE)
depth_range_m
# remove - not needed since all records taken at 0m (sea level)

# 4) Lat (degrees north) & Long (degrees east)
lat_range <- range(monterey_sst$latitude, na.rm = TRUE)
long_range <- range(monterey_sst$longitude, na.rm = TRUE)
lat_range
long_range
# plot below - checked bounding area, explore more below

# 5) sst (Celsius) - check range
sst_summary <- summary(monterey_sst$sst, na.rm = TRUE)
sst_summary
# remove - NA values

# 6) sst anomaly (Celsius) - check range
anom_summary <- summary(monterey_sst$anom, na.rm = TRUE)
anom_summary
# ok
```

## Explore Further - Lat/Long

```{r}
# Investigate -- quick map of lat/long points for sst and ebird sightings

# please note, you need the bird_sf dataframe from the eBird_Cleaning.Rmd

# Convert to sf
noaa_m_sf <- monterey_sst %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# World/US map background
world <- ne_countries(scale = "medium", returnclass = "sf")

# Plot from Bodega Bay to Monterey Bay
ggplot() +
  geom_sf(data = world, fill = "gray96", color = "gray80") +
  geom_sf(data = bird_sf, aes(color = "Shearwater Sightings"), size = 1) +
  geom_sf(data = noaa_m_sf, aes(color = "SST Record Locations"), size = 1.5) +
  scale_color_manual(
    name = "Data Source",
    values = c(
      "Shearwater Sightings" = "lightblue",
      "SST Record Locations" = "black"
    )
  ) +
  coord_sf(xlim = c(-123.9, -120.5), ylim = c(34.8, 37.1)) +
  theme_minimal() +
  labs(
    title = "Monterey County Shearwater Sightings and SST Recordings") +
  theme(plot.caption = element_text(hjust = 0, size = 10, color = "darkgray"),
        axis.text = element_text(size = 11, color = "gray"),
        legend.title = element_blank(),
        legend.position = c(0.999, 0.08),
        legend.justification = c(1, 0),
        legend.box.background = element_rect(
          fill = "white",
          color = "black",
          linewidth = 0.5),
    panel.background = element_rect(fill = "aliceblue", color = NA))
# thoughts: it looks like the sst data points cover a good amount of the area 
# where shearwaters were sighted off the Monterey coast.
```

# Changes to Data Based on Integrity Analyses

```{r}
# Clean --

# Note: I already changed longitude from 0–360 system to the −180–180 system 
# when I created the master dataframe.

# Combined cleaning pipelines
sst_clean <- monterey_sst %>%
  
  # 1) time/depth - remove unnecessary columns
  select(-time, -datetime, -depth) %>%
  
  # 2) sst - remove missing SST values
  filter(!is.na(sst)) %>%
  
  # reorder columns with date first
  select(date, everything())

# Check
head(sst_clean)

# save the data as an Excel file
library(writexl)
write_xlsx(sst_clean, "sst_clean.xlsx")
```
