---
title: "Data Preperations"
author: "Danielle D."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

# Analysis Introduction

# Loading Work Space & Set Up

```{r}
# packages
library(readxl) # read in xl files
library(ggplot2) # plotting
library(tidyverse) # data manipulation
library(lubridate) # date handling

# read in cleaned data
bird_df <- read_excel("data/clean_bird.xlsx")
sst_df <- read_excel("data/clean_sst.xlsx")

# check that all of the data types are correct, no changes needed
str(bird_df)
str(sst_df)
```

# Aggregate SST by Month

```{r}
sst_monthly <- sst_df %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d"),
         year = year(date),
         month = month(date),
         month_date = floor_date(date, "month")) %>%
         # month_date will make it easier to plot and join data later
  group_by(year, month, month_date) %>%
  summarise(sst_monthly_avg = mean(sst),
            sd_sst = sd(sst),
            anom_monthly_avg = mean(anom),
            sd_anom = sd(anom),
            n = n(),
            .groups = "drop")

sst_monthly # check data
```

# Daily Max Bird Counts

Observers can count the same birds in a flock, so I first took the maximum
number of birds counted per day. I did not average the counts per month because
we'd lose pulses in counts.

```{r}
# summarize and group my species and county per day
bird_daily <- bird_df %>%
  mutate(date = as.Date(date, format = "%Y-%m-%d"),
         year = year(date),
         month = month(date),
         month_date = floor_date(date, "month")) %>%
  group_by(date, year, month, month_date, county, common_name) %>%
  summarise(daily_max_count = max(observation_count, na.rm = TRUE),
            .groups = "drop")

# check
head(bird_daily)
nrow(bird_df)
nrow(bird_daily)

# note: I can decide later to keep more columns from bird_df here if needed.
```

# Join Aggregated SST and Bird Counts

```{r}
bird_sst <- bird_daily %>%
  left_join(sst_monthly, by = c("month_date", "year", "month"))

bird_sst # check

# save the data as an Excel file
library(writexl)
write_xlsx(sst_clean, "data/bird_sst.xlsx")
```

# Preliminary Plots

1) Raw counts over time by species and county

```{r}
ggplot(bird_sst, aes(x = date, y = daily_max_count, color = common_name)) +
  geom_point(alpha = 0.5) +
  scale_y_log10() +
  facet_wrap(~county) +
  labs(title = "Daily Max Counts Over Time", x = "Date", y = "Count (log scale)") +
  theme_minimal()
```

2) Count distribution by species

```{r}
ggplot(bird_sst, aes(x = common_name, y = daily_max_count, fill = common_name)) +
  geom_boxplot() +
  scale_y_log10() +
  facet_wrap(~county) +
  labs(title = "Count Distribution by Species", x = NULL, y = "Count (log scale)") +
  theme_minimal()
``` 

3) Seasonal arrival pattern — counts by month

```{r}
ggplot(bird_sst, aes(x = month, y = daily_max_count, color = common_name)) +
  geom_point(alpha = 0.4) +
  geom_smooth(se = FALSE) +
  scale_y_log10() +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  facet_wrap(~county) +
  labs(title = "Seasonal Count Patterns", x = "Month", y = "Count (log scale)") +
  theme_minimal()
```

4) SST over time (Monterey only)

```{r}
sst_monthly %>%
  ggplot(aes(x = month_date, y = sst_monthly_avg)) +
  geom_line(color = "steelblue") +
  geom_ribbon(aes(ymin = sst_monthly_avg - sd_sst,
                  ymax = sst_monthly_avg + sd_sst), alpha = 0.2, fill = "steelblue") +
  labs(title = "Monthly SST - Monterey", x = "Date", y = "SST (°C)") +
  theme_minimal()
```

5) SST vs count scatter (Monterey only)

```{r}
bird_sst %>%
  filter(county == "Monterey") %>%
  ggplot(aes(x = sst_monthly_avg, y = daily_max_count, color = common_name)) +
  geom_point(alpha = 0.5) +
  scale_y_log10() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "SST vs Count - Monterey", x = "SST (°C)", y = "Count (log scale)") +
  theme_minimal()
```

6) Check seasonality of SST temps

```{r}
bird_sst %>%
  filter(county == "Monterey", !is.na(sst_monthly_avg)) %>%
  mutate(month_label = factor(month.abb[month], levels = month.abb)) %>%
  ggplot(aes(x = sst_monthly_avg, y = daily_max_count, color = month_label)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_y_log10() +
  scale_color_viridis_d(option = "turbo") +
  labs(title = "SST vs Count colored by Month - Monterey",
       x = "SST (°C)", y = "Count (log scale)", color = "Month") +
  theme_minimal()
```

Preliminary from the graphs:

- PFSH have a more northerly distribution tendency than SOSH, with more PFSH
  appearing in Sonoma county than SOSH.
- Shearwater counts are increasing with sst, however this is shown to be due to
  their seasonal arrival in spring and summer, so higher SST does not 
  necessarily drive higher counts.

