---
title: "Data Cleaning – Shearwaters Climate Project"
author: "Danielle D."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
description: >
  This notebook details the data cleaning workflow for the Shearwaters Climate project,
  including preprocessing steps, variable transformations, and reproducibility notes.
---

# Data Cleaning

## Overview

This document walks through the full data cleaning process for the Shearwaters Climate project. 

```{r}
# packages used
library(readr) # read eBird data in
library(tidyverse) 
library(janitor) # used to standardize column names
library(lubridate)  # for easier date handling
library(ggplot2) 
```


## eBird Data

This data included records of sooty shearwater and pink-footed shearwater sightings in the Monterey and Bodega Bay areas (Monterey county and Sonoma county). The sighting data includes the date, time, count, and occasionally other breeding data or comments.

Put in the read me:
Bird observation data sourced from the eBird Basic Dataset (Cornell Lab of Ornithology). Used under the Creative Commons Attribution-NonCommercial license.

eBird Basic Dataset. Version: [year]. Cornell Lab of Ornithology, Ithaca, New York. https://ebird.org/data/download.

```{r}
# Load eBird Data --

# Data was pulled from Monterey, Santa Cruz, and Bodega counties for the dates 
# January 1st, 2015 - November 30th, 2025.
  # Bird codes:
  # sooty shearwater (SOSH)
  # pink-footed shearwater (PFSH)

# ----- pfsh.monterey

pfsh.monterey <- read_delim(
  "/Users/danielledevincenzi/Desktop/Learning R/eBird data/ebd_US-CA-053_pifshe_201501_202511_smp_relOct-2025/ebd_US-CA-053_pifshe_201501_202511_smp_relOct-2025.txt",
  delim = "\t",
  show_col_types = FALSE
)

# standardize column names with janitor package 
pfsh.monterey <- pfsh.monterey %>%
  janitor::clean_names()

# ----- sosh.monterey

sosh.monterey <- read_delim(
  "/Users/danielledevincenzi/Desktop/Learning R/eBird data/ebd_US-CA-053_sooshe_201501_202511_smp_relNov-2025/ebd_US-CA-053_sooshe_201501_202511_smp_relNov-2025.txt",
  delim = "\t",
  show_col_types = FALSE
)

sosh.monterey <- sosh.monterey %>%
  janitor::clean_names()

# ----- pfsh.sonoma

pfsh.sonoma <- read_delim(
  "/Users/danielledevincenzi/Desktop/Learning R/eBird data/ebd_US-CA-097_pifshe_smp_relNov-2025/ebd_US-CA-097_pifshe_smp_relNov-2025.txt",
  delim = "\t",
  show_col_types = FALSE
)

pfsh.sonoma <- pfsh.sonoma %>%
  janitor::clean_names()

# ----- sosh.sonoma

sosh.sonoma <- read_delim(
  "/Users/danielledevincenzi/Desktop/Learning R/eBird data/ebd_US-CA-097_sooshe_201501_202511_smp_relNov-2025/ebd_US-CA-097_sooshe_201501_202511_smp_relNov-2025.txt",
  delim = "\t",
  show_col_types = FALSE
)

sosh.sonoma <- sosh.sonoma %>%
  janitor::clean_names()

# ----- combine eBird data

combined <- rbind(pfsh.monterey, sosh.monterey, pfsh.sonoma, sosh.sonoma)

```

```{r}
# Cleaning eBird data --

# ----- pull columns of interest to create master
bird_master <- combined %>%
  select(state_code, county, locality, locality_id, latitude, longitude, 
         last_edited_date, common_name, observation_count, observation_type, 
         duration_minutes, effort_distance_km, number_observers, observer_id, group_identifier)

bird_master

# ----- organize dates and filter out the range

# Make sure date column is in Date or POSIXct format
bird_master$last_edited_date <- as.Date(bird_master$last_edited_date)

# filter data by exactly 10 years of data (Jan. 1, 2015 - Dec. 31, 2024)
bird_master <- bird_master %>%
  filter(last_edited_date >= "2015-01-01" & last_edited_date <= "2024-12-31")

# check dates were updated
date_range <- range(bird_master$last_edited_date, na.rm = TRUE)
date_range

# ----- remove odd entries

# Make sure counts are numeric
bird_master <- bird_master %>%
  mutate(observation_count = as.numeric(observation_count))

# Check max and min
max_count <- max(bird_master$observation_count, na.rm = TRUE)
min_count <- min(bird_master$observation_count, na.rm = TRUE)

max_count
min_count

# Flag all unusually high counts, e.g., over 100,000
extreme_counts <- bird_master %>%
  filter(observation_count > 100000) %>%
  select(last_edited_date, county, observation_count, locality, observer_id, group_identifier)

# View them
extreme_counts

# check group names
unique(extreme_counts$group_identifier)

# only three groups make up all of these outliers, I would exclude. Likely extra 0 added, need to look into high counts below 10000 as well. Graph again once 100k+ entries are removed. Organize this cleaning section, don't overwrite dataframe, may want to retunr to unclean data. Graph as yhou go and include justification for removal.
```

```{r}
# Graph eBird counts by date

# ----- make observation count column numeric
bird.master <- bird.master %>%
  mutate(observation_count = as.numeric(observation_count))

# ----- summarize counts per day per county
bird.master <- bird.master %>%
  group_by(last_edited_date, county) %>%
  summarize(total_count = sum(observation_count, na.rm = TRUE)) %>%
  ungroup()

# ----- plot
ggplot(bird.master, aes(x = last_edited_date, y = total_count, color = county)) +
  geom_line() +              # lines per county
  geom_point(alpha = 0.5) +  # optional: points for each day
  labs(
    x = "Date",
    y = "Total Observations",
    color = "County",
    title = "Shearwater Observations by County Over Time"
  ) +
  theme_minimal()
```

## NOAA Data

This data includes temperature, wind, sea surface temperature (sst), upwelling index, and enso index for Monterey and Bodega Bays.

Data sourced from NOAA [dataset name], accessed on [date]. NOAA data are in the public domain.

```{r}
# weekly average? daily - too noisy
# identify anomalies
```
