---
title: "Data Cleaning – Shearwaters Climate Project"
author: "Danielle D."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
description: >
  This notebook details the data cleaning workflow for the Shearwaters Climate project,
  including preprocessing steps, variable transformations, and reproducibility notes.
---

# Data Cleaning

## Overview

This document walks through the full data cleaning process for the Shearwater project. 

## NOAA Data

This data includes...
  - Sea surface temperature (sst) : ties to marine heatwaves, upwelling strength, and forage fish distribtuion
  - Wind: strong northwest wings indicate cold, productive waters
  - Upwelling Index (Bakun Index): productivity
  - ENSO indices: large scale climate patterns that affect local productivity

for Monterey and Bodega Bays.

Data sourced from NOAA [dataset name], accessed on [date]. NOAA data are in the public domain.

```{r}
library(readr)
one <- read.csv("noaa ata/ocldb1764961090.2001649.GLD.csv", header = F)
two <- read.csv("noaa data/ocldb1764961090.2001649.MRB.csv", header = F)
three <- read.csv("noaa data/ocldb1764961090.2001649.OSD.csv", header = TRUE)
four <- read.csv("noaa data/ocldb1764961090.2001649.APB.csv", header = TRUE)
five <- read.csv("noaa data/ocldb1764961090.2001649.PFL.csv", header = TRUE)

# weekly average? daily - too noisy
# identify anomalies
```

## eBird Data

This data included records of sooty shearwater and pink-footed shearwater sightings in the Monterey and Bodega Bay areas. The sighting data includes the date, time, count, and occasionally other breeding data or comments.

Put in the read me:
Bird observation data sourced from the eBird Basic Dataset (Cornell Lab of Ornithology). Used under the Creative Commons Attribution-NonCommercial license.

eBird Basic Dataset. Version: [year]. Cornell Lab of Ornithology, Ithaca, New York. https://ebird.org/data/download.

```{r}
# sooty shearwater (SOSH)

sosh.santa.cruz

sosh.monterey

sosh.bodega

# pink-footed shearwater (PFSH)

pfsh.santa.cruz

pfsh.monterey

sosh.bodega
```

