---
title: "Data Cleaning – Shearwaters Climate Project"
author: "Danielle D."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

# Data Cleaning

## Overview

This document walks through the full data cleaning process for the Shearwater project. 

This document walks through the full data cleaning process for the Shearwater
Climate project. 

```{r}
# packages used
library(readr) # to read eBird data
library(tidyverse) # for data manipulation
library(janitor) # used to standardize column names
library(lubridate) # for easier date handling
library(ggplot2) # for plotting
```

<<<<<<< HEAD
This data includes...
  - Sea surface temperature (sst) : ties to marine heatwaves, upwelling strength, and forage fish distribtuion
  - Wind: strong northwest wings indicate cold, productive waters
  - Upwelling Index (Bakun Index): productivity
  - ENSO indices: large scale climate patterns that affect local productivity

for Monterey and Bodega Bays.
=======
## eBird Data
 
This data included records of sooty shearwater (SOSH) and pink-footed shearwater (PFSH) sightings in the Monterey and Bodega Bay areas (Monterey county and Sonoma county). The sighting data includes the date, time, and count of individuals.

Bird observation data sourced from the eBird Basic Dataset (Cornell Lab of Ornithology). Used under the Creative Commons Attribution-NonCommercial license.

  eBird Basic Dataset. Version: 2025. Cornell Lab of Ornithology, Ithaca, New York.                    https://ebird.org/data/download.

```{r}
# Load eBird Data --

# Data was pulled from Monterey, Santa Cruz, and Bodega counties for the dates 
# January 1st, 2015 - November 30th, 2025.

pfsh.monterey <- read_delim(
  "/Users/danielledevincenzi/Desktop/Learning R/eBird data/ebd_US-CA-053_pifshe_201501_202511_smp_relOct-2025/ebd_US-CA-053_pifshe_201501_202511_smp_relOct-2025.txt",
  delim = "\t",
  show_col_types = FALSE
)

sosh.monterey <- read_delim(
  "/Users/danielledevincenzi/Desktop/Learning R/eBird data/ebd_US-CA-053_sooshe_201501_202511_smp_relNov-2025/ebd_US-CA-053_sooshe_201501_202511_smp_relNov-2025.txt",
  delim = "\t",
  show_col_types = FALSE
)

pfsh.sonoma <- read_delim(
  "/Users/danielledevincenzi/Desktop/Learning R/eBird data/ebd_US-CA-097_pifshe_smp_relNov-2025/ebd_US-CA-097_pifshe_smp_relNov-2025.txt",
  delim = "\t",
  show_col_types = FALSE
)

sosh.sonoma <- read_delim(
  "/Users/danielledevincenzi/Desktop/Learning R/eBird data/ebd_US-CA-097_sooshe_201501_202511_smp_relNov-2025/ebd_US-CA-097_sooshe_201501_202511_smp_relNov-2025.txt",
  delim = "\t",
  show_col_types = FALSE
)
```

```{r}
# Combine eBird data --

# first, combine the above dataframes
combined <- rbind(pfsh.monterey, sosh.monterey, pfsh.sonoma, sosh.sonoma)

# standardize column names with janitor package 
combined <- combined %>%
  janitor::clean_names()

# check parsing issue from above chunk - not a big deal, don't need this column
problems(combined)

# pull columns of interest to create master
bird_master <- combined %>%
  select(state_code, county, locality, locality_id, latitude, longitude, 
         last_edited_date, common_name, observation_count, observation_type, 
         duration_minutes, effort_distance_km, number_observers, observer_id, group_identifier)

# bird_master # check
```

```{r}
# Clean eBird data - Dates

# make sure date column is in Date or POSIXct format
bird_master$last_edited_date <- as.Date(bird_master$last_edited_date)

# filter data by exactly 10 years of data (Jan. 1, 2015 - Dec. 31, 2024)
bird_master <- bird_master %>%
  filter(last_edited_date >= "2015-01-01" & last_edited_date <= "2024-12-31")

# check dates were updated
date_range <- range(bird_master$last_edited_date, na.rm = TRUE)
date_range
```


```{r}
library(readr)

ebird_data <- read_delim(
  "eBird data/ebd_US-CA-053_pifshe_201501_202511_smp_relOct-2025/ebd_US-CA-053_pifshe_201501_202511_smp_relOct-2025.txt",
  delim = "\t",
  show_col_types = FALSE
)

library(janitor)

ebird_data <- ebird_data %>%
  janitor::clean_names()

ebird_data

library(dplyr)
library(lubridate)  # for easier date handling

# Make sure date column is in Date or POSIXct format
ebird_data$last_edited_date <- as.Date(ebird_data$last_edited_date)

# 1. Unique bird species
unique_species <- unique(ebird_data$common_name)
length(unique_species)   # How many species in total
head(unique_species, 20) # See first 20 species

# 2. Date range
date_range <- range(ebird_data$last_edited_date, na.rm = TRUE)
date_range

# --------------

ebird_data2 <- read_delim(
  "eBird data/ebd_US-CA-087_pifshe_201501_202511_smp_relOct-2025/ebd_US-CA-087_pifshe_201501_202511_smp_relOct-2025_sampling.txt",
  delim = "\t",
  show_col_types = FALSE
)

ebird_data2 <- ebird_data2 %>%
  janitor::clean_names()

ebird_data2

# Make sure date column is in Date or POSIXct format
ebird_data2$last_edited_date <- as.Date(ebird_data2$last_edited_date)

# 1. Unique bird species
unique_species <- unique(ebird_data2$common_name)
length(unique_species)   # How many species in total
head(unique_species, 20) # See first 20 species

# 2. Date range
date_range <- range(ebird_data2$last_edited_date, na.rm = TRUE)
date_range

# --------------

ebird_data3 <- read_delim(
  "eBird data/ebd_US-CA-087_pifshe_201501_202511_smp_relOct-2025/ebd_US-CA-087_pifshe_201501_202511_smp_relOct-2025_sampling.txt",
  delim = "\t",
  show_col_types = FALSE
)

ebird_data3 <- ebird_data2 %>%
  janitor::clean_names()

ebird_data3

# Make sure date column is in Date or POSIXct format
ebird_data3$last_edited_date <- as.Date(ebird_data3$last_edited_date)

# 1. Unique bird species
unique_species <- unique(ebird_data3$common_name)
length(unique_species)   # How many species in total
head(unique_species, 20) # See first 20 species

# 2. Date range
date_range <- range(ebird_data3$last_edited_date, na.rm = TRUE)
date_range

# to do next...
# figure out what data I have here (county, dates, species) - what do I still need to request?
# combine all eBird data, get rid of columns you don't need

# PFSH and SOSH data for Santa Cruz and Monterey counties, Bodega bay county too

# Clean eBird data - Counts

# make sure counts are numeric
bird_master <- bird_master %>%
  mutate(observation_count = as.numeric(observation_count))

# check max and min
max_count <- max(bird_master$observation_count, na.rm = TRUE)
min_count <- min(bird_master$observation_count, na.rm = TRUE)

max_count # this seems too high
min_count # none go below 0

# flag all unusually high counts, e.g., over 100,000
extreme_counts <- bird_master %>%
  filter(observation_count > 200000) %>%
  select(last_edited_date, county, observation_count, locality, observer_id, group_identifier)

# view them
extreme_counts

# check group names
unique(extreme_counts$group_identifier)

# Only three groups make up all of the extreme counts. 

# remove extreme counts
bird_master_clean <- bird_master %>%
  filter(observation_count <= 200000 | is.na(observation_count))
```
# Everything below this is in progress...

```{r}
# Graph counts

# ----------------------
# gather the max counts per day per county 
# note: can't sum the counts daily because observers count the same birds
bird_master_day <- bird_master_clean %>%
  group_by(last_edited_date, county) %>%
  summarize(
    daily_max = max(observation_count, na.rm = TRUE),
    .groups = "drop"
  )

# plot - bar and points
ggplot(bird_master_day, aes(x = last_edited_date, y = daily_max, color = county)) +
  geom_line() +              # lines per county
  geom_point(alpha = 0.5) +  # optional: points for each day
  labs(
    x = "Date",
    y = "Total Observations",
    color = "County",
    title = "Shearwater Observations Over Time"
  ) +
  theme_minimal()

# ----------------------
# heat map
heat <- bird_master_clean %>%
  mutate(
    week = week(last_edited_date),
    month = month(last_edited_date, label = TRUE)
  ) %>%
  group_by(county, year, week) %>%  # or month
  summarise(weekly_max = max(max_count, na.rm = TRUE), .groups = "drop")

ggplot(heat, aes(x = week, y = factor(year), fill = weekly_max)) +
  geom_tile() +
  facet_wrap(~county) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    x = "Week of Year",
    y = "Year",
    fill = "Max Count",
    title = "Weekly Max Shearwater Counts Heatmap"
  ) +
  theme_minimal()

# ----------------------
# Prepare data for box and density plots
shearwater <- bird_master_clean %>%
  mutate(
    date = as_date(last_edited_date),
    year = year(date),
    month = month(date, label = TRUE),
    day_of_year = yday(date)
  ) %>%
  group_by(county, date, year, day_of_year) %>%  # <-- group by county + date
  summarize(
    daily_max = max(observation_count, na.rm = TRUE),
    .groups = "drop"
  )

# Boxplots per year
ggplot(shearwater, aes(x = factor(year), y = daily_max, fill = county)) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 1) +
  labs(
    x = "Year",
    y = "Daily Max Count",
    fill = "County",
    title = "Distribution of Daily Max Counts per Year"
  ) +
  theme_minimal()

# Density plots per year (seasonality)
ggplot(shearwater, aes(x = day_of_year, y = ..density.., color = county, fill = county)) +
  geom_density(alpha = 0.3) +
  labs(
    x = "Day of Year",
    y = "Density of Max Counts",
    title = "Seasonal Density of Shearwater Daily Max Counts"
  ) +
  theme_minimal()

```

```{r}
# Brainstorming...

# Fix graphs above, choose most representative.

# Organize this cleaning section, don't overwrite dataframes, may want to return to unclean data. Graph as you go and include justification for removal.

# I'm noticing no observations in Monterey pre 2015... could this be no birds or something else, like eBird surged in popularity in 2020 with covid? Need to look into this further.

# Think about the goal, shiny app visualization of what? seasonality? interested in changes over time with climate variables... cool ways to visualize data... 
```

## NOAA Data

This data includes temperature, wind, sea surface temperature (sst), upwelling index, and enso index for Monterey and Bodega Bays.

Data sourced from NOAA [dataset name], accessed on [date]. NOAA data are in the public domain.

https://www.ncei.noaa.gov/products/optimum-interpolation-sst
https://www.cpc.ncep.noaa.gov
https://coastwatch.pfeg.noaa.gov/data.html
https://coastwatch.pfeg.noaa.gov/index.html

```{r}
# weekly average? daily - too noisy
# identify anomalies
```
