---
title: "NOAA Data Cleaning Script"
author: "Danielle D."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

# Introduction

This data includes temperature, wind, sea surface temperature (sst), upwelling index, and enso index for Monterey and Bodega Bays.

Data sourced from NOAA [dataset name], accessed on [date]. NOAA data are in the public domain.

https://www.ncei.noaa.gov/products/optimum-interpolation-sst
https://www.cpc.ncep.noaa.gov
https://coastwatch.pfeg.noaa.gov/data.html
https://coastwatch.pfeg.noaa.gov/index.html

This data includes...
  - Sea surface temperature (sst) : ties to marine heatwaves, upwelling strength, and forage fish distribution
  - Wind: strong northwest wings indicate cold, productive waters
  - Upwelling Index (Bakun Index): productivity
  - ENSO indices: large scale climate patterns that affect local productivity

for Monterey and Bodega Bays.

```{r}
# weekly average? daily - too noisy
# identify anomalies
```

# Loading Workspace & Set Up

```{r}
# packages
# packages
library(readr) # read data
library(janitor) # standardize column names
library(tidyverse) # data manipulation
library(lubridate) # date handling
library(ggplot2) # plotting
```


# Integrity Analysis

# Changes to Data Based on Integrity Analyses

