---
title: "NOAA Data Cleaning Script"
author: "Danielle D."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

# Introduction

Open source sea surface temperature data (sst) in Celsius for Monterey and 
Sonoma counties.

Data sourced from NOAA OISST-V2-AVHRR Daily (1/4 degree By time, depth, latitude, longitude), accessed on February 25, 2026. NOAA data are in the public domain.
	
Huang, Boyin; Liu, Chunying; Banzon, Viva F.; Freeman, Eric; Graham, Garrett; Hankins, William; Smith, Thomas M.; Zhang, Huai-Min. (2020): NOAA 0.25-degree Daily Optimum Interpolation Sea Surface Temperature (OISST), Version 2.1. [indicate subset used]. NOAA National Centers for Environmental Information. https://doi.org/10.25921/RE9P-PT57. Accessed [February 2026].

SST has ties to marine heatwaves, upwelling strength, and forage fish 
distribution, all of which could affect arrival for shearwaters. 

# Loading Workspace & Set Up

```{r}
# packages
library(readr) # read data
library(janitor) # standardize column names
library(tidyverse) # data manipulation
library(lubridate) # date handling
library(ggplot2) # plotting
```

## Reading in the Data

```{r}
# Load --

# use "here" to manage raw files housed out of project (too large for GitHub)
library(here)

#  define path to raw SST data (outside project)
raw_data_dir <- "~/Desktop/R/Shearwater raw data/"

#  load CSV for Sonoma County SST
sonoma_sst <- read_csv(
  file.path(
    raw_data_dir,
    "S_ncdc_oisst_v2_avhrr_by_time_zlev_lat_lon_df91_44aa_5324.csv"
  ),
  show_col_types = FALSE
)

# load CSV for Monterey County SST
monterey_sst <- read_csv(
  file.path(
    raw_data_dir,
    "M_ncdc_oisst_v2_avhrr_by_time_zlev_lat_lon_9cc6_587b_5479.csv"
  ),
  show_col_types = FALSE
)

# check
head(monterey_sst)
head(sonoma_sst)
```

## Creating Master Dataframes

```{r}
# Clean --

sonoma_sst <- sonoma_sst %>%
  slice(-1) %>% # remove first row of data, additional header data
  mutate( 
    datetime = ymd_hms(time, tz = "UTC"),   # convert character to POSIXct (date)
    date = as_date(datetime),               # extract date
    time = format(datetime, "%H:%M:%S")     # extract time as string
  ) %>%
  mutate(sst = as.numeric(sst))             # make sst numeric

monterey_sst <- monterey_sst %>%            # clean in the same way
  slice(-1) %>% 
  mutate(
    datetime = ymd_hms(time, tz = "UTC"),  
    date = as_date(datetime),               
    time = format(datetime, "%H:%M:%S")     
  ) %>%
    mutate(sst = as.numeric(sst))

# check
head(monterey_sst)
head(sonoma_sst)
```

# Integrity Analysis

```{r}
# Explore --

# 1. Date/Time - check range (UTC)
date_range_m <- range(monterey_sst$date)
date_range_s <- range(sonoma_sst$date)
print("Monterey")
date_range_m
print("Sonoma")
date_range_s
# date range is 2/28/2020 to 12/26/2025 for both sites

time_range_m <- range(monterey_sst$time)
time_range_s <- range(sonoma_sst$time)
print("Monterey")
time_range_m
print("Sonoma")
time_range_s
# All times are 12:00 PM, standardized
# remove - date/time and time columns

# 2. Depth - (m)
depth_range_m <- range(monterey_sst$depth, na.rm = TRUE)
depth_range_s <- range(sonoma_sst$depth, na.rm = TRUE)
print("Monterey")
depth_range_m
print("Sonoma")
depth_range_s
# I wasn't interested in depth in my data, all 0.0m depth
# remove - depth column

# 3. Lat (degrees north) & Long (degrees east)
# map below
# 

# 4. SST - (Celsius)
sst_summary_m <- summary(monterey_sst$sst, na.rm = TRUE)
sst_summary_s <- summary(sonoma_sst$sst, na.rm = TRUE)
print("Monterey")
sst_summary_m
print("Sonoma")
sst_summary_s
# clean NAs, fix sonoma data, other data looks ok
```

## Explore Further - Lat/Long

```{r}
# overlay eBird counts on map!!
```


# Changes to Data Based on Integrity Analyses

# Explore Clean Data
