---
title: "NOAA Data Cleaning Script"
author: "Danielle D."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---

# Introduction

Open source sea surface temperature data (sst) in Celsius for Monterey and 
Sonoma counties.

Data sourced from NOAA OISST-V2-AVHRR Daily (1/4 degree By time, depth, latitude, longitude), accessed on February 25, 2026. NOAA data are in the public domain.
	
Huang, Boyin; Liu, Chunying; Banzon, Viva F.; Freeman, Eric; Graham, Garrett; Hankins, William; Smith, Thomas M.; Zhang, Huai-Min. (2020): NOAA 0.25-degree Daily Optimum Interpolation Sea Surface Temperature (OISST), Version 2.1. [indicate subset used]. NOAA National Centers for Environmental Information. https://doi.org/10.25921/RE9P-PT57. Accessed [February 2026].

SST has ties to marine heatwaves, upwelling strength, and forage fish 
distribution, all of which could affect arrival for shearwaters. 

# Loading Workspace & Set Up

```{r}
# packages
library(readr) # read data
library(janitor) # standardize column names
library(tidyverse) # data manipulation
library(lubridate) # date handling
library(ggplot2) # plotting
```

```{r}
library(here)
library(readr) # read_csv

#  Define path to your raw SST data (outside project)
raw_data_dir <- "~/Desktop/R/Shearwater raw data/"

#  Load CSV for Sonoma County SST
sonoma_sst <- read_csv(
  file.path(
    raw_data_dir,
    "S_ncdc_oisst_v2_avhrr_by_time_zlev_lat_lon_df91_44aa_5324.csv"   # replace with your CSV filename
  ),
  show_col_types = FALSE
)

# ⃣ Load CSV for Monterey County SST
monterey_sst <- read_csv(
  file.path(
    raw_data_dir,
    "M_ncdc_oisst_v2_avhrr_by_time_zlev_lat_lon_9cc6_587b_5479.csv"  # replace with your CSV filename
  ),
  show_col_types = FALSE
)

head(monterey_sst)
head(sonoma_sst)
```

# Integrity Analysis

# Changes to Data Based on Integrity Analyses

# Explore Clean Data
